From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-28 Pi/paper/ufin.tex
Citation updates
Use uniform style for references
Add signatures for funext/ua
Add a line about equalities

%   \note{We use the rig structure of $\UFin$ in~\Cref{subsec:rig} to interpret $\PiLang$.}

@@

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a hierarchy of (univalent) universes $\UU_0 : \UU_1 : \ldots$ (though we
will just write $\UU$ since we mostly use one universe), and a few Higher Inductive Types (HITs) for truncations and
quotients. We write $\dfun*{x:A}{B(x)}$ or simply $(x:A) \to B(x)$ for dependent function types, and $\dsum*{x:A}{B(x)}$
or simply $(x:A) \times B(x)$ for dependent sum types.

@@

A homotopy between functions $f, g : A \to B$, written $f \htpy g$, is given by pointwise equality between them,
$\dfun*{x:A}{f(x) \id_{B} g(x)}$. The identity type for functions is equivalent to homotopies between them ${(f \id_{A
\to B} g)} \eqv {(f \htpy g)}$, by $\term{funext}$ and $\term{happly}$. An equivalence between types $A \eqv B$ is given
by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
\idfunc_{A}$ (also see~\cref{prop:three-chars-equiv}), and this is equivalent to the identity type for the universe, $(A
\id_{\UU} B) \eqv (A \eqv B)$, which is the univalence principle.

Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
universe, such as $A \to \UU$, which is an $A$-indexed family of groupoids. For a type family $P : A \to \UU$ and a
point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation, named
$\term{transport}/\term{tr}$, lifts paths in the indexing type to functions between fibers.
\begin{gather*}
  \term{ap}_{f} : \dfun*{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
  \qquad
  \term{transport}_{P} : \dfun*{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\end{gather*}

@@

\begin{toappendix}
  Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
  \[
    \term{lift}(u,p) : (x , u) \id_{\dsum*{x:A}{P(x)}} (y , \transport{P}{p}{u})
  \]

>>>>

%   \note{We use the rig structure of $\UFin$ in~\cref{subsec:rig} to interpret $\PiLang$.}

@@

We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
intensional Martin-L\"{o}f Type Theory, with a hierarchy of (univalent) universes $\UU_0 : \UU_1 : \ldots$ (though we
will just write $\UU$ since we mostly use one universe), and a few Higher Inductive Types (HITs) for truncations and
quotients. We write $\dfun*{x:A}{B(x)}$ or simply $(x:A) \to B(x)$ for dependent function types, and $\dsum*{x:A}{B(x)}$
or simply $(x:A) \times B(x)$ for dependent sum types. We use $\defeq$ for definitions, $\equiv$ for definitional
equalities, and $\id$ for the identity type.

@@

A homotopy between functions $f, g : A \to B$, written $f \htpy g$, is given by pointwise equality between them,
$\dfun*{x:A}{f(x) \id_{B} g(x)}$. The identity type for functions is equivalent to homotopies between them ${(f \id_{A
\to B} g)} \eqv {(f \htpy g)}$, by $\term{funext}$ and $\term{happly}$. An equivalence between types $A \eqv B$ is given
by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
\idfunc_{A}$ (also see~\cref{prop:three-chars-equiv}), and this is equivalent to the identity type for the universe, $(A
\id_{\UU} B) \eqv (A \eqv B)$, which is the univalence principle.
\begin{gather*}
  \begin{aligned}
    \term{happly} & : (f \id g) \to (f \htpy g) \\
    \term{funext} & : (f \htpy g) \to (f \id g)
  \end{aligned}
  \qquad
  \begin{aligned}
    \term{idtoeqv} & : (A \id B) \to (A \eqv B) \\
    \term{ua}      & : (A \eqv B) \to (A \id B)
  \end{aligned}
\end{gather*}

Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
universe, such as $A \to \UU$, which is an $A$-indexed family of groupoids. For a type family $P : A \to \UU$ and a
point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation, named
$\term{transport}$, lifts paths in the indexing type to functions between fibers.
\begin{gather*}
  \term{ap}_{f} : \dfun*{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
  \qquad
  \term{transport}_{P} : \dfun*{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
\end{gather*}

@@

\begin{toappendix}
  Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
  \[
    \term{lift}(u,p) : (x , u) \id_{\dsum*{x:A}{P(x)}} (y , \transport{P}{p}{u})
  \]
  shown in~\cref{fig:lift}.
